A family of accessibility measures

### 2024-12-13 paper/manuscript.qmd
5-slide trb pres draft

The aim of this paper's focus is on the proportionality constant. We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values with clear meanings. To address this matter, we need not reinvent the wheel: instead we refer to the spatial interaction model as developed by @wilson1971. In essence, in what follows, we reunite the accessibility indicators literature with the spatial interaction literature as developed by @wilson1971 based on entropy maximization concepts. Although Wilson's approach is based on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants: including the one that eluded @stewartDemographicGravitationEvidence1948 and has been ignored in almost all subsequent accessibility research.

@@

$$
\sum_i\sum_j T_{ij} c_{ij} = C
$${#eq-constraint0_5-gravitymodel}

@@

The entropy maximation derivation is demonstrated in detail elsewhere (e.g., @ortuzar_2011_modelling, @wilsonSTATISTICALTHEORYSPATIAL1967). The basis of the approach is to accept that there are macro, meso, and micro states. In the case of trips being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of indiivudal trip makers, and the macro state is the entire study region. Using these assumptions, we aim to find a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel). From this approach, the gravity model can be interpreted to say that given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level). There is then a most probable distribution of trips between zones (meso-states) that stasis all the macro-state constraints.

@@

- The unconstrained case: neither @eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold;
- The total constrained case: only @eq-constraint0-gravitymodel hold
- The singly-constrained case: either @eq-constraint1-gravitymodel holds (origin-constrained) <!-- production--> or @eq-constraint2-gravitymodel holds (destination-constrained)<!-- attraction-->; 
- The doubly-constrained case: both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously <!--production-attraction-->

@@

For the doubly-constrained case, $A_i$ and $B_j$ must be solved together along with $K$. For the singly-constrained cases, origin-constrained case replaces $B_j$ with 1 for all $j$ and $A_i$ is @eq-bf-A-gravitymodel, and the the destination-constrained case replaces $A_i$ with 1 for all $i$ and $B_j$ is @eq-bf-B-gravitymodel. For both singly-constrained cases, $K$ also holds. For the total constrained case, $A_i$ and $B_j$ are set to 1 and $K$ is the defining constraint. And for the unconstrained case, in a way, all other cases (the total constrained, singly-constrained and doubly-constrained cases) are simply constrained variations.

@@

Imagine if the interaction variable was in fact _potential_ interaction, and the units were trips but opportunities. We don't have to change much! Referring to the preceding section, the entropy maximizing method can be used to derive an equivalent family of measures, but for the case of accessibility where the _interaction_ variable is $V_{ij}$, represent _potential interaction_, i.e., the number of potential reachable opportunities. $T_{ij}$ in @eq-phys-gravity-model can be replaced with $V_{ij}$ and presented as follows:

$$
V_{ij} = k W_i^{(1)} W_j^{(2)} f(c_{ij})
$${#eq-access}

Where $V_{ij}$ consists of the same variables, however $V_{ij}$ explicitly measures _potential interaction_ as a unit of number of potentially reachable opportunities and the single proportionality constant $k$ ensures the total sum of $V_{ij}$ represents the total potential reachable opportunities in the data.

@@

$$
\sum_i\sum_j V_{ij} = V
$${#eq-constraint0-access}

$$
\sum_i\sum_j V_{ij} c_{ij} = C
$${#eq-constraint0_5-access}
\noindent Where the sum of the travel cost of each potentially reached opportunity in the region is equal to a total travel cost $C$.

$$
\sum_j V_{ij} = W_i^{(1)}
$${#eq-constraint1-access}
\noindent Where sum of all potentially reachable opportunities for all $j$ zones in the region is equal to the production weight at each $i$ in the region.

$$
\sum_i V_{ij} = W_j^{(2)}
$${#eq-constraint2-access}
\noindent Where sum of all potentially reachable opportunities for all $i$ zones in the region is equal to the attraction weight at each $j$ in the region.

@@

## Unconstrained accessibility 
(no proportionality constant, k=1, no constraints hold - hard to interpret):
$$
S_{ij} = P_i O_j f(c_{ij})
$$
In practice, $S_i$ is calculated for each $i$ so:
$$
S_{i} = P_i \sum_j O_j f(c_{ij})
$$
These numbers are proportional to the accessibility (i.e., higher or lower), but are not scaled to number opportunities (jobs).

@@

This needs a balancing factor, $k$ is replaced with $K$:
$$
S_{ij}' = K P_i O_j f(c_{ij})\\
S_{i}' = K P_i \sum_j O_j f(c_{ij})
$$

@@

Similarly, singly-constrained can be either origin-constrained $V_i''^{(1)}$ or destination-constrained $V_i''^{(2)}$. For the origin- and destination- constrained versions, $k$ is replaced with $A_i$ and $B_j$ respectively:
$$
V_{ij}''^{(1)} = A_i P_i O_j f(c_{ij})\\
V_{ij}''^{(2)} = B_j P_i O_j f(c_{ij})\\

V_{i}''^{(1)} = A_i P_i \sum_j O_j f(c_{ij})\\
V_{i}''^{(2)} = P_i \sum_j B_j O_j f(c_{ij})\\

@@

\noindent Where $A_i$ and $B_j$ are:
$$
A_i = \frac{1}{\sum_j O_j f(c_{ij})}\\
B_j = \frac{1}{\sum_i P_i f(c_{ij})}\\
$$
For the origin-constrained, $V_{i}''^{(1)}$ are the number of opportunities (like jobs) that can be reached at $i$, and $V_{i}''^{(2)}$ are the number of opportunities (like people) that can be reached at $i$

In terms of origin-constrained -- spatial availability is explicitly so [@soukhovIntroducingSpatialAvailability2023;@soukhovMultimodalSpatialAvailability2024]. Other literature also demonstrates this constraint implicitly through its analytical formulation; for instance the Intervening Opportunities model [@stoufferInterveningOpportunitiesTheory1940; @stoufferINTERVENINGO PPORTUNITIESCOMPETING1960] "axiomatically" @weibull_axiomatic_1976, competition consideration but as a per population rate [@shen1998], FCA methods in general [].

Regarding destination-constrained, has it's origins are in retail models. e.g., decision model for shops to optimally spatially locate to maximise their potential consumer interaction. Starting with reilly's law of gravitational retail (1931), formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965.

>>>>

This paper's focus is on the proportionality constant. We wish to demonstrate the importance of transiting from proportionality to equality in @eq-accessibility-general, in order to balance the units of accessibility into more interpretable values with clear meanings. To address this matter, we need not reinvent the wheel: instead we refer to the spatial interaction model as developed by @wilson1971. In essence, in what follows, we reunite the accessibility indicators literature with the spatial interaction literature as developed by @wilson1971 based on entropy maximization concepts. Although Wilson's approach is based on a different conceptual foundation than the old reference to Newtonian gravity, the work succeeded at identifying the steps from proportionality to equality to yield variations of proportionality constants: including the one that eluded @stewartDemographicGravitationEvidence1948 and has been ignored in almost all subsequent accessibility research.

@@

<!-- $$ -->
<!-- \sum_i\sum_j T_{ij} c_{ij} = C -->
<!-- $${#eq-constraint0_5-gravitymodel} -->

@@

The entropy maximation derivation is demonstrated in detail elsewhere (e.g., @ortuzar_2011_modelling, @wilsonSTATISTICALTHEORYSPATIAL1967). The basis of the approach is to accept that there are macro, meso, and micro states. In the case of trips being the interaction variable, a micro state is an individual trip maker, the meso state is the statistically representative zone representing a collection of individual trip makers, and the macro state is the entire study region. Using these assumptions, we aim to find a set of $T_{ij}$ that maximizes the number of micro states in each meso state subject to some sensible macro-state constraints (@eq-constraint0_5-gravitymodel, @eq-constraint1-gravitymodel, and @eq-constraint2-gravitymodel). From this approach, the gravity model can be interpreted to say that given the total number of individual trips (micro level) statistically represented at origin and destination zones (meso level), the cost of travelling (micro level, represented at the meso level), and the fixed total cost of travelling in the region (number of trips and cost of travel at the macro level). There is then a most probable distribution of trips between zones (meso-states) that stasis all the macro-state constraints.

@@

- The unconstrained case: neither @eq-constraint0-gravitymodel, @eq-constraint1-gravitymodel, or @eq-constraint2-gravitymodel hold;
- The total constrained case: only @eq-constraint0-gravitymodel hold
- The singly-constrained case: either @eq-constraint1-gravitymodel holds (origin-constrained) <!-- production--> or @eq-constraint2-gravitymodel holds (destination-constrained)<!-- attraction-->; 
- The doubly-constrained case: both @eq-constraint1-gravitymodel and @eq-constraint2-gravitymodel hold simultaneously <!--production-attraction-->

@@

For the doubly-constrained case, $A_i$ and $B_j$ must be solved together along with $K$. For the singly-constrained cases, origin-constrained case replaces $B_j$ with 1 for all $j$ and $A_i$ is @eq-bf-A-gravitymodel, and the the destination-constrained case replaces $A_i$ with 1 for all $i$ and $B_j$ is @eq-bf-B-gravitymodel. For both singly-constrained cases, $K$ also holds. For the total constrained case, $A_i$ and $B_j$ are set to 1 and $K$ is the defining constraint. And for the unconstrained case, all other cases (the total constrained, singly-constrained and doubly-constrained cases) are simply constrained variations.

@@

Instead of the interaction variable being in the unit of trips, it is in the unit of potential interaction with opportunities. This focus is the preoccupation of accessibility measures. The same analytical framework to define the balancing factors $k$ can be used for the family of accessibility measures as in the family of gravity models, however with different interpretation. In this section, we explicitly define these cases.

In this context, accessibility is our _interaction_ variable of choice, so $T_{ij}$ in @eq-phys-gravity-model is replaced with $V_{ij}$. As the interaction variable represents _potential_, i.e., the number of potential reachable opportunities, as opposed to trips:

<!-- W_i^{(1)} enters through k.. it is a quality of the destination -->

$$
V_{ij} = k W_j^{(2)} f(c_{ij})
$${#eq-access}

Where $V_{ij}$ consists of the same variables, however $V_{ij}$ explicitly measures _potential interaction_ as a unit of number of potentially reachable opportunities and the single proportionality constant $k$ ensures the total sum of $V_{ij}$ represents the total potential reachable opportunities in the data. Accessibility is also most often presented as a locational measure, the value of $V_i$ is of most interest, e.g., the number of potentially reachable opportunities at a zone.

@@

$$
\sum_i\sum_j V_{ij} = V
$${#eq-constraint0-access}
\noindent Where the sum of all potentially reached opportunities in the region is equal to a total opportunities in the region $V$.

$$
\sum_j V_{ij} = W_i^{(1)} = P_i
$${#eq-constraint1-access}
\noindent Where sum of all potentially reachable opportunities for all $j$ zones in the region is equal to the production weight at each $i$ in the region.

$$
\sum_i V_{ij} = W_j^{(2)} = O_j
$${#eq-constraint2-access}
\noindent Where sum of all potentially reachable opportunities for all $i$ zones in the region is equal to the attraction weight at each $j$ in the region.

@@

## Unconstrained accessibility

- No proportionality constant, $k$ is set to 1, so no constraints hold, making the interpretation of the units meaningless so the value's magnitude can only be interpreted:
$$
V_{ij} = k O_j f(c_{ij})
$$

- In practice, $V_i$ is also calculated for each $i$. When thinking about the interpretation of the value, how many opportunities can be visited for each $i$, the $P_i$ is also set to 1:
$$
V_{i} = \sum_j O_j f(c_{ij})
$$
These numbers are proportional to the accessibility (i.e., higher or lower), but are not scaled to number opportunities (jobs).

@@

This needs a balancing factor, $k$ is replaced with $K$:
$$
V_{ij}' = K P_i O_j f(c_{ij})\\
V_{i}' = K P_i \sum_j O_j f(c_{ij})
$$

@@

Similarly, singly-constrained can be either origin-constrained $V_i''^{(1)}$ or destination-constrained $V_i''^{(2)}$. For the origin- and destination- constrained versions, $k$ is replaced with $A_i$ and $B_j$ respectively:
$$
V_{ij}''^{(1)} = A_i O_j f(c_{ij})\\
V_{ij}''^{(2)} = B_j O_j f(c_{ij})\\

@@

\noindent Where $A_i$ and $B_j$ are:
$$
A_i = \frac{1}{\sum_j O_j f(c_{ij})}\\
B_j = \frac{1}{\sum_i P_i f(c_{ij})}\\
$$

In terms of origin-constrained -- ..? .

Regarding destination-constrained, has it's origins are in retail models. e.g., decision model for shops to optimally spatially locate to maximise their potential consumer interaction. Starting with reilly's law of gravitational retail (1931), formalized later in @huffDefiningEstimatingTrading1964 and @lakshmananRetailMarketPotential1965. Spatial availability is explicitly also destination-constrained [@soukhovIntroducingSpatialAvailability2023;@soukhovMultimodalSpatialAvailability2024]. Other literature also demonstrates this constraint implicitly through its analytical formulation; for instance the Intervening Opportunities model [@stoufferInterveningOpportunitiesTheory1940; @stoufferINTERVENINGO PPORTUNITIESCOMPETING1960] "axiomatically" @weibull_axiomatic_1976, competition consideration but as a per population rate [@shen1998], FCA methods in general []
